Challenges in Developing Great QMC Software

### 2022-12-27 MCQMC2022ChallengesQMCSoftwareArticle.tex
minor updates

@@

\title*{Challenges in Developing Great MCQMC Software}
\authorrunning{S.-C.\ T.\ Choi et al}
\author{Sou-Cheng Terrya Choi \and Yuhan Ding \and Fred J. Hickernell \and Aleksei G. Sorokin}
\institute{Sou-Cheng Terrya Choi \at Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{schoi32@hawk.iit.edu}
\and
Fred J. Hickernell \at Center for Interdisciplinary Scientific Computation and \\
Department of Applied Mathematics, Illinois Institute of Technology \\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{hickernell@iit.edu}
\and Yuhan Ding \at
\and 
Aleksei G. Sorokin \at
Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{asorokin@hawk.iit.edu}}

@@

\abstract{}

@@

Monte Carlo and their often more efficient variant, quasi-Monte Carlo (QMC) methods, have developed since the middle of the last century.  With the explosion in computational science, there is a need for great Monte Carlo and Quasi-Monte Carlo (MCQMC) software to realize the value in MCQMC methods. Based on our observations of what software has been developed to date, and what potential is not yet realized, we propose some criteria for developing great MCQMC software and some practical steps forward.  Our focus is primarily on QMC software.

	\begin{itemize}
	\item Correct, e.g., not omit the zeroth point in a low discrepancy sequence \cite{Owe22a, scipySobol2020a}

	\item  Complete---contain the components or  easily access components in other libraries to solve real, complex problems

	\item  Accessible---downloadable; tutorials, demos, discussion forums, etc.\ for (new) users; written in a language that potential users speak; provide a consistent user interface

	\item Efficient---in terms of compute time and memory

	\item Current---include the latest and best algorithms

	\item  Sustainable---have a sufficient user base and developer community for updates and maintenance

	\item Scalable---take advantage of advanced computer architectures for speed and to solve large problems

	\begin{itemize}
	\item \cite{LEc2017a} provides a history of random number generation
	\item Notorious RANDU from the 1960s \cite{RANDU}, which fails the spectral test
	\item VEGAS \cite{Lep78a, Lep21a} importance sampling Monte Carlo algorithm favored by physicists
	\item ACM low discrepancy point generators \cite{BraFox88,BraFoxNie92,HonHic00a}
	\item FinDer \cite{PasTra95,FinDer} and BRODA \cite{BRODA20a} targeting quantitative finance
	\item Korobov cubature and scrambled Sobol' generators in NAG \cite{NAG27} for decades		
	\item Scrambled Sobol' and Halton generators in MATLAB \cite{MAT9.13} since 2008, and fixed a few years later
	\item BUGS \cite{BUGSBook, BUGSweb} and Stan \cite{STAN} for Markov Chain Monte Carlo (MCMC)
	\item LatMRG \cite{LEcCou97}, RNGStreams \cite{LEcEtal02},  SSJ \cite{LEc2002a,SSJ}, and LatNetBuilder \cite{LatNet} 
	\item Low discrepancy generators \cite{FriKel02,FriKelweb}, SamplePack \cite{SamplePack}, \cite{GruWeb},  and MatBuilder \cite{paulin2022} 
	\item Sobol' direction numbers \cite{JoeKuo03,JoeKuo08,SobolDirection}
	\item Fast CBC, Magic Point Shop, QMC4PDE and other code since 2004  at \cite{NuyWeb} 
	\item Multi-level software \cite{GilesSoft,GilesQSoft} 
	\item Data-driven error bounds and stopping criteria in GAIL \cite{ChoEtal21a} and QMCPy \cite{QMCPy2020a, ChoEtal22a}  
	\item Uncertainty quantification libraries Dakota \cite{DakotaUsersManual}, UQTk \cite{DebEtal04,UQTk}, and MUQ \cite{MUQ}  have some basic level low discrepancy sampling
	\item QMC framework in Julia since 2019 \cite{QMCJulia} by Robbe and others 
	\item Scrambled Sobol' in SciPy \cite{virtanen2020scipy} and PyTorch \cite{paszke2019pytorch} since several years ago
	\item TensorFlow QMC framework \cite{tfqfQMC2021a} since a year or so ago
\end{itemize}

We have stand-alone algorithms, small libraries, and pieces of larger libraries

@@

>>>>

%\input FJHDef.tex
\newcommand{\AGSComment}[1]{{\color{red} #1}}

@@

\title*{Challenges in Developing Great QMC Software}
\authorrunning{S.-C.\ T.\ Choi et al}
\author{Sou-Cheng Terrya Choi \and Yuhan Ding \and Fred J. Hickernell \and Aleksei G. Sorokin}
\institute{Sou-Cheng Terrya Choi \at Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{schoi32@hawk.iit.edu}
\and
Fred J. Hickernell \at Center for Interdisciplinary Scientific Computation and \\
Department of Applied Mathematics, Illinois Institute of Technology \\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{hickernell@iit.edu}
\and Yuhan Ding \at
\and 
Aleksei G. Sorokin \at
Department of Applied Mathematics, Illinois Institute of Technology,\\ RE 220, 10 W.\ 32$^{\text{nd}}$ St., Chicago, IL 60616 \email{asorokin@hawk.iit.edu}}

@@

\abstract{
Monte Carlo and their often more efficient variant, quasi-Monte Carlo (QMC) methods, have developed since the middle of the last century.  With the explosion in computational science, there is a need for great Monte Carlo and Quasi-Monte Carlo (MCQMC) software to realize the value in MCQMC methods. Based on our observations of what software has been developed to date, and what potential is not yet realized, we propose some criteria for developing great MCQMC software and some practical steps forward.  Our focus is primarily on QMC software.
}

\newpage

@@

Modern programming languages enable a simple Monte Carlo method approximating $\mu := \mathbb{E}[f(\boldsymbol{X})]$, $\boldsymbol{X} \sim \mathcal{U}[0,1]^d$ to be implemented in three lines of code:
\begin{enumerate}
    \item draw $n$ independent identically distributed (IID) standard uniform nodes 
    $$\boldsymbol{x}_1,\dots,\boldsymbol{x}_n \sim \mathcal{U}[0,1]^d,$$
    \item evaluate the integrand $f$ at these nodes to obtain $$y_i := f(\boldsymbol{x}_i), \qquad i=1,\dots,n$$
    \item estimate the true mean $\mu$ by the sample mean
    \begin{equation}
        \hat{\mu} := \frac{1}{n} \sum_{i=1}^n y_i.
        \label{eq:mc_average}
    \end{equation}  
\end{enumerate}
However, practitioners may want to utilize advanced sampling methods such as antithetic sampling, importance sampling, or low discrepancy sampling in the case of Quasi-Monte Carlo (QMC) methods. Users may also wish to modify the integrand to accommodate taking the expectation with respect to a non-standard-uniform random variable or to utilize control variates. The above enhancements may require taking the weighted average in place of \eqref{eq:mc_average} to maintain an unbiased estimate. Practitioners may also be interested in determining $n$ adaptively so the approximation is within some error threshold from the true mean $\mu$.

Developing software to implement such enhanced MC strategies poses a number of challenges. First, a development team must determine the value of the software to the users. We discuss why great (Q)MC software in necessary in Section \ref{sec:why_we_need_software}. Second, a team must determine high level objectives for the software; the tenets of great software are described in Section \ref{sec:tenets_of_great_software}. Third, an architecture for the  (Q)MC software should be determined. We propose a modularization of the (Q)MC pipeline in Section \ref{sec:qmc_software_architecture}. Next, a team must choose which programming languages, hardware environments, and distribution platforms they wish to support. We discuss the nuance and challenges in these decisions in Section \ref{sec:language_library_environment}. Finally, a project which hopes to stand the test of time must draw attention and support from the broader (Q)MC community. To this end, we encourage collaborative community development and describe its' best practices in  Section \ref{sec:encouraging_collaborative_developement}.

\section{Why Develop Software} \label{sec:why_we_need_software}

In 2021, the US Department of Energy held a ``Workshop on the Science of Scientific-Software Development and Use'' \cite{Her19a,ASCR-SSSDU}. This workshop not only discussed diagnostics and treatments for the challenges of developing great scientific software, but also emphasized the importance of implementing theoretical advancements into well written, accessible software libraries. This section describes both the practical and theoretical importance of developing scientific software.

\subsection{Theory to Software}

The chief purpose of most scientific software developments is to make theoretical advancements available to practitioners. Developing theory into practical software leads to algorithms which often require additional considerations. For instance, the numerical stability of a method only becomes relevant when considering how to implement a method onto a finite precision computer. Taking the time to consider such implications and determine relevant algorithm parameters alleviates these burdens form the user. Moreover, the availability of such algorithms enables practitioners with a variety of applications to quickly solve problems.

\subsection{Software to Theory}

An often overlooked benefit of scientific software  is the feedback garnered for future theoretical development. A few motivating examples are
\begin{itemize}
    \item the early application of QMC to high dimensional option pricing and 
    \item the early application of neural networks to problems in computer vision.
\end{itemize}
The unexpected performance in these examples motivated theoretical developments to justify their success. Moreover, the availability of high quality software will often attract practitioners to a certain methodology over competitors, and in turn may inspire further theoretical development.

\section{Tenets of Great Software} \label{sec:tenets_of_great_software}

In this section we lay out common aspirations of scientific software. While some objectives are expected by practitioners, e.g. correctly written software, others may better be characterized as features that are pursued as a project matures, e.g. supporting scalable computation. The following paragraphs highlight aspirations with more important, expected behaviours coming earlier in the discussion.

Software is expected to be \emph{correct}. A practitioner generating a digital sequence expects the routine to produce theoretically accurate points. For example, the QMC community has helped developers ensure implementations of low discrepancy sequence generators include the zeroth point \cite{Owe22a, scipySobol2020a} and  have correct randomization routines. %, and disallow thinning / leaping: taking every $k^\text{th}$ value in the sequence.

Algorithms should be implemented in an \emph{efficient} manner. Both the compute time and memory requirements should match theoretical developments which in turn should be as close to the state of the art as possible. As a classical example, it is expected that an implementation of the discrete Fourier transform for $n$ points has complexity $\mathcal{O}(n\log n)$.

Software should be \emph{accessible} in terms of installation, presentation, navigation, exemplification, and communication. A library should be straightforward to install from a common distribution platform, consistent in user interface, intuitive to navigate to modules and documentation, comprehensible in its' examples, and supportive in user engagement. Software accessibility considerations are further discussed in Section \ref{sec:language_library_environment}.

Libraries should be \emph{sustainable} for future development. The goal should be to have a sufficient user base and developer community for updates and maintenance. Similar to accessibility, sustainability is often overlooked for aspirations discussed later in this section. However, we contend that a sub-optimal but user-friendly library with community engagement may be more impactful than an optimal library which lacks support. We promote collaborative, community driven development further in Section \ref{sec:encouraging_collaborative_developement}.

Software should be \emph{integrated} into the environment's scientific software ecosystem. More simply, libraries should play nicely together and utilize the mature developments of other teams. A simple example in \AGSComment{CITE} shows a straightforward integration between the QMC package QMCPy \cite{QMCPy2020a} and the PDE package FEniCS/Dolfin \cite{LoggEtal_10_2012}, both in Python.

Routines should be \emph{scalable} to advanced computer architectures. The widespread availability of on demand computing with multi-core and multi-GPU machines encourages software that can take advantage of parallel computation. Scalable algorithms lend themselves to a larger class or problems and provide practitioners the opportunity to more quickly test ideas.

\section{(Quasi-)Monte Carlo Software Architecture} \label{sec:qmc_software_architecture}

An important step in the development of any software is to determine its' architecture. How should components interact with each other? Which aspects should be modular? Object oriented? Procedural? Answers to these questions form the blueprint of software design and guide implementation.

This section proposes an object oriented design for (Q)MC software. The authors have implemented this architecture into the QMCPy \cite{QMCPy2020a} Python package which is further detailed in the MCQMC 2020 proceedings article \AGSComment{CITE}. While this is not the only possible architecture for (Q)MC problems, we hope the high level object definitions facilitate easy collaboration and integration for research across the (Q)MC community.

The standard (Q)MC problem is to approximate the true mean $\mu := \mathbb{E}[g(\boldsymbol{T})]$ where $\boldsymbol{T}$ is some $d$ dimensional random variable. For nice $\boldsymbol{T}$, one may perform a change of variables to write $\mu = \mathbb{E}[f(\boldsymbol{X})]$ where $\boldsymbol{X} \sim \mathcal{U}[0,1]^d$. For example, if $\boldsymbol{T} \sim \mathcal{N}(\boldsymbol{0},\boldsymbol{I})$, then one may set $f(\boldsymbol{x}) := g(\boldsymbol{\Phi}^{-1}(\boldsymbol{x}))$ for $\boldsymbol{x} \in [0,1]^d$ where $\boldsymbol{\Phi}^{-1}$ is the inverse CDF of a standard normal distribution taken element wise. One may then approximate $\mu$ by the $n$ sample average $\hat{\mu} = 1/n \sum_{i=1}^n f(\boldsymbol{x}_i)$ where $\{\boldsymbol{x}_i\}_{i=1}^n$ are chosen so their empirical distribution is ``close'' to the uniform distribution. The approximation error $\varepsilon = \lvert \mu - \hat{\mu} \rvert$ is $\mathcal{O}(n^{-1/2})$ when $\{\boldsymbol{x}_i\}_{i=1}^n$ are chosen to be IID standard uniform as in the case of simple Monte Carlo (MC) and almost $\mathcal{O}(n^{-1})$ when $\{\boldsymbol{x}_i\}_{i=1}^n$ are chosen from a low discrepancy sequence as done for Quasi-Monte Carlo (QMC). Instead of selecting the number of samples $n$ and then determining the error $\varepsilon$, it is often desirable to select $\varepsilon$ and then adaptively determine $n$. From this view, $n$ is $\mathcal{O}(\varepsilon^{-2})$ for MC and almost $\mathcal{O}(\varepsilon^{-1})$ for QMC. A more detailed account of MC and QMC can be found in \AGSComment{CITE}.

The (Q)MC pipeline above can be decomposed into $4$ main components:  
\begin{description}
    \item[\textbf{Generator:}] producing $\{\boldsymbol{x}_i\}_{i=1}^n$ whose empirical distribution is ``close'' to the standard uniform distribution in $d$ dimensions for any $n \geq 1$.  May produce IID or low discrepancy sequences depending on the problem. 
    \item[\textbf{Measure:}] that defines the distribution of $\boldsymbol{T}$ e.g., uniform, Gaussian, Lebesgue. This definition facilitates choosing a change of variables to determine $f$. 
    \item[\textbf{Integrand:}] $g$, which defines the original integral. When provided knowledge of the true measure, the integrand can also evaluate $f$ via the change of variables.  
    \item[\textbf{Stopping Criterion:}] based on a data-driven error bound, which determines how large $n$ should be to ensure that $\varepsilon$ is less than some user-provided error threshold.
\end{description}
A variety of QMC materials and the their categorizations into the above framework are described in the following list.
\begin{itemize}
    %\item \cite{LEc2017a} provides a history of random number generation
    %\item Notorious RANDU from the 1960s \cite{RANDU}, which fails the spectral test
    %\item VEGAS \cite{Lep78a, Lep21a} importance sampling Monte Carlo algorithm favored by physicists
    \item ACM low discrepancy point generators \cite{BraFox88,BraFoxNie92,HonHic00a}
    \item FinDer \cite{PasTra95,FinDer} and BRODA \cite{BRODA20a} targeting quantitative finance
    \item Korobov cubature and scrambled Sobol' generators in NAG \cite{NAG27} for decades	
    \item Scrambled Sobol' and Halton generators in MATLAB \cite{MAT9.13} since 2008, and fixed a few years later
    %\item BUGS \cite{BUGSBook, BUGSweb} and Stan \cite{STAN} for Markov Chain Monte Carlo (MCMC)
    \item LatMRG \cite{LEcCou97}, RNGStreams \cite{LEcEtal02},  SSJ \cite{LEc2002a,SSJ}, and LatNetBuilder \cite{LatNet} 
    \item Low discrepancy generators \cite{FriKel02,FriKelweb}, SamplePack \cite{SamplePack}, \cite{GruWeb},  and MatBuilder \cite{paulin2022} 
    \item Sobol' direction numbers \cite{JoeKuo03,JoeKuo08,SobolDirection}
    \item Fast CBC, Magic Point Shop, QMC4PDE and other code since 2004  at \cite{NuyWeb} 
    \item Multi-level software \cite{GilesSoft,GilesQSoft} 
    \item Data-driven error bounds and stopping criteria in GAIL \cite{ChoEtal21a} and QMCPy \cite{QMCPy2020a, ChoEtal22a}  
    \item Uncertainty quantification libraries Dakota \cite{DakotaUsersManual}, UQTk \cite{DebEtal04,UQTk}, and MUQ \cite{MUQ}  have some basic level low discrepancy sampling
    \item QMC framework in Julia since 2019 \cite{QMCJulia} by Robbe and others 
    \item Scrambled Sobol' in SciPy \cite{virtanen2020scipy} and PyTorch \cite{paszke2019pytorch} since several years ago
    \item TensorFlow QMC framework \cite{tfqfQMC2021a} since a year or so ago
    \item CUBA \cite{CUBA}
\end{itemize}

\section{Choosing a Environment} \label{sec:language_library_environment}

This section discusses the challenges in choosing environments for development and distribution. The development environment includes the choice of which languages, architectures, testing methods, and documentation methods to support. The deployment environment will often be influenced by the choice of programming languages.

The choice of programming language should be aligned with the communities the team wishes to reach with their software. The choice should also account for developer familiarity and ability to quickly prototype ideas. A team should align their choice of language to balance the desire to have user-friendly prototyping code versus the desire to have high performance computing (HPC) ``ninja'' code written for production setting. To the second point, it is also important to consider which language will best facilitate the parallel and GPU computations the team wishes to support.

Many programming languages come bundled with their own preferred distribution systems. Python has PyPI, R has CRAN, Julia has Pkg. However, many projects support multiple languages and have complex dependency requirements. In such cases, it can be helpful to containerize your application using a service like Docker.

\section{Encouraging Collaborative Development} \label{sec:encouraging_collaborative_developement}

We now discuss ways a team may encourage collaborative development while touching on a few best practices. As discussed earlier, collaborative development promotes greater community engagement, ecosystem integration, and software longevity.

The authors have found the Git version control system to be an invaluable for software development. Git hosts such as GitHub or Bitbucket may offer teams additional features such as tracking issues, managing discussion threads, and setting up automated workflows for jobs such as testing and documentation compilation.

Accessibility is key to community engagment. It is paramount that your package be easy to install, well documented, and tested regularly. Aside from static communications, a team should also be engaged in addressing users bugs and be attentive to collaborating with proposed code changes.

It is also important for a development team to make a clear path for collaborators to contributors to the software. Templates for issues and pull requests are a good place to start. Contributors should also be encouraged to write robust tests for their updates and highlight their new features in a demo for greater accessibility.

@@

\AGSComment{
\tableofcontents
\section*{TODO}
\begin{itemize}
    \item Flesh out Section \ref{sec:encouraging_collaborative_developement}
    \item QMC Software Architecture section: better way then description of components and list of softwares? 
    \item For Section \ref{sec:tenets_of_great_software}, the slides also include 
    \begin{itemize}
        \item  Complete---contain the components or  easily access components in other libraries to solve real, complex problems
        \item Current---include the latest and best algorithms
    \end{itemize}
    which I think fall under the umbrella of software being \emph{integrated} with the community.
    \item update QMCPy citation \cite{QMCPy2020a}
    \item figures?
    \item I have left out or been breif about material from the following slides: QMCPy + FEniCS/Dolfin,  Building a Developer Team for Your Library
\end{itemize}
}
